Research log

### 2024-01-17 docs/paper.Rmd
incorporating stereotypicality analysis

---
title             : "Pretending not to know: Pretense behavior reveals the limits of self-simulation"
shorttitle        : "Pretending not to know"

@@

To this end, we examine pretense in a game setting. Using an online version of the games Battleship and Hangman (in which players seek to uncover the locations of enemy ships or the identity of a word), participants played a 'non-pretend' (normal) version of the game, as well as a 'pretend' version where they were given complete information about the hidden ships / target word but were instructed to behave as if they didn't have this information. Overall, we find a highly impressive capacity for pretending not to know, with pretenders mirroring broad patterns and subtle features of real players' decisions and decision times. However, we also find that pretense was characterized by pervasive biases and limitations, although these were undetected by human observers.

@@

```{r design, echo=FALSE, fig.cap="Experimental Design in Exp. 1 (left) and 2 (right). In non-pretend games, players revealed ships by guessing cells in a grid (A) or revealed a word by guessing letters (E). In pretend games, we marked ship locations with a cross (B) and revealed the target word from the start (F), but asked players to play as if they didn't have this information. In half-games (C and G), players continued the game from a given state, with the content of some cells and the position of some letters already revealed. Lastly, players watched replays of the games of previous players and guessed which were pretend games (D and H).", out.width = '100%'}
knitr::include_graphics("figures/methods.png")
```

@@

## Pretense limitations

Though impressive, the capacity for simulating a state of ignorance was not perfect. First, despite showing the same qualitative effects, decision time patterns in Battleship pretend games (but not Hangman pretend games) were systematically more pronounced relative to non-pretend games, as if participants were acting based on an exaggerated mental model of their own counterfactual behavior. Furthermore, the within-participant sequential guess entropy, a measure of the inverse predictability of decision sequences across different games of the same player, was significantly reduced in pretend games (Fig. \@ref(fig:rigidity)A; see [online supplementary materials](https://matanmazor.github.io/ignorance/docs/exp.-1-battleship-exploratory-results.html#within-participant-click-entropy)). More specifically, Hangman pretenders were more likely to guess letters that appear frequently in English words (E, T, A, etc.) irrespective of the game state, compared to genuine players (Fig. \@ref(fig:rigidity)B). This suggests that in their attempt to behave as if they didn't know the true state of the game, pretenders had an increased tendency to follow rigid heuristics and rules, ignoring useful information as a result.

This limitation on incorporating evidence into the (simulated) decision-making process was especially evident in Hangman half-games, where players completed the game from a half-completed state. When asked to reveal the hidden fruit "\_A_A_A", `r round(100*E4.first_click_half_games %>% filter(test_part=='nonpretend_half_game' & word %in% c('banana', 'papaya'))%>%pull(bananahand)%>%mean())`% of the non-pretenders guessed one of the letters 'B' or 'N' (Fig. \@ref(fig:rigidity)C, yellow bars in left column). Among pretenders who knew that the hidden word was BANANA, this preference was reduced to `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word== 'banana')%>%pull(bananahand)%>%mean())`%. Importantly, half of the pretenders were given different information: they were told that the hidden word was the less prototypical fruit PAPAYA. Although good pretenders should simulate their behavior had they not known this information, only `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='papaya')%>%pull(bananahand)%>%mean())`% selected the letters 'B' or 'N', revealing that many pretenders were unable to predict that the first fruit that would have come to their minds was BANANA, not PAPAYA (Fig. \@ref(fig:rigidity)C, yellow bars in right column). A similar pattern was observed for the prototypical body part word HA(ND) and its surprising counterpart HA(IR): when playing normally, `r round(100*E4.first_click_half_games %>% filter(test_part=='nonpretend_half_game' & word %in% c('hand', 'hair'))%>%pull(bananahand)%>%mean())`% of the players selected letters that are consistent with the prototypical option HAND. This figure was `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='hand')%>%pull(bananahand)%>%mean())`% among pretenders for whom the target word was HAND, in contrast to only `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='hair')%>%pull(bananahand)%>%mean())`% among pretenders for whom the target word was HAIR (Fig. \@ref(fig:rigidity)C, blue bars).

```{r rigidity, echo=FALSE, message= F, warning=F, fig.cap="Limitations on flexible decision-making when pretending. A: sequential guess entropy, a measure of the (inverse) predictability of individual players' guesses as a function of click number and guess number. In both Battleship and Hangman, sequential guess entropy increased with click number, and was overall lower in pretend games. B: letter frequency of Hangman guesses: the median rank frequency per participant in pretend and non-pretend games, with reference lines for the expected rank frequency for a random agent, and for a “greedy” agent that maximizes the probability of a hit in each step. C: first letter guesses in Hangman half-games, as a function of pretense condition and target word. Letters appear in alphabetical order. letters that appear in the target word are marked in yellow (BANANA), red (PAPAYA), blue (HAND) and magenta (HAIR). For reference, the overall distribution of letter guesses across all games and conditions is given below. ", out.width = '100%'}
knitr::include_graphics("figures/rigidity.png")
```

@@

Despite these limitations, pretenders' behavior resembles real players' behavior closely enough that they were able to fool other participants into believing they were playing for real. After performing pretend and non-pretend games, participants were presented with game replays of previous players and took on the new role of being a 'judge' who had to determine who was pretending and who was playing for real. In both games, pretense detection was at chance (Battleship: `r E2.judge_df%>%group_by(subj_id)%>%summarise(accuracy=sum(correct)*20)%>%pull(accuracy)%>%mean()%>%round()`%, Hangman: `r round(E4.judge_summary_df$accuracy%>%mean()*100)`%). This is in line with previous findings of near-chance accuracy in lie detection [@bond2006]. Moreover, we find no sign of a correlation between pretense quality (measured as players' ability to trick judges into thinking they were not pretending) and pretense detection ability (measured as proportion correct; Battleship: `r apa_print(cor.test(E2.performance_df$judge_rate,E2.performance_df$pretend_rate, method='spearman'))$estimate`, Hangman: `r apa_print(cor.test(E4.performance_df$pretend_rate,E4.performance_df$judge_rate,method='spearman'))$estimate`), indicating that pretense and pretense detection rely on at least partly different cognitive processes.

@@

An alternative interpretation of our results is that instead of simulating a counterfactual knowledge state, participants actively suppressed or ignored the revealed game state such that their entire cognitive machinery was available to play the game. This would not require self-simulation, only a capacity to intentionally 'unsee', or forget, relevant evidence. While we cannot fully rule out this interpretation, we think it is unlikely to explain our players' successful pretense, for at least four reasons. First, we tried to make such suppression as hard as possible, by presenting the game solution on top of the game board for the entire duration of pretend games, and by having participants type the target word before pretend Hangman games. Second, suppressing thoughts on demand is notoriously difficult, and often has an opposite, positive effect on the suppressed content [@wegner1987paradoxical]. Third, when asked how they had performed the task in a debrief question, the responses of a significant majority of participants were aligned with self-simulation or rule-following (see [exploratory analysis](https://matanmazor.github.io/ignorance/docs/exp.-1-battleship-exploratory-results.html#suppression-or-simulation)). Finally, response time patterns were exaggerated in pretend Battleship relative to non-pretend Battleship games (see Fig. \@ref(fig:time)A): a finding that is consistent with the simulation of a simplified self-model, but is hard to explain if participants were merely suppressing their current knowledge.

@@

The research complied with all relevant ethical regulations and was approved by the Institutional Review Board of Johns Hopkins University. `E2.df$subj_id%>%unique()%>%length()` Participants were recruited via Prolific (prolific.co) and gave their informed consent prior to their participation. They were selected based on their acceptance rate (\>95%) and for being native English speakers. The entire experiment took approximately 20 minutes to complete. Participants' pay was equivalent to an hourly wage of 9.50 USD, in addition to a bonus payment (``` r``E2.bonus$bonus%>%min() ``` - `r E2.bonus$bonus%>%max()` USD, mean = `r E2.bonus$bonus%>%mean()%>%printnum()`).

>>>>

---
title             : "Pretending not to know reveals a powerful capacity for self-simulation"
shorttitle        : "Pretending not to know"

@@

To this end, we examine pretense in a game setting. Using an online version of the games Battleship and Hangman (in which players seek to uncover the locations of enemy ships or the identity of a word), participants played a 'non-pretend' (normal) version of the game, as well as a 'pretend' version where they were given complete information about the hidden ships / target word but were instructed to behave as if they didn't have this information. Overall, we find a highly impressive capacity for pretending not to know, with pretenders mirroring broad patterns and subtle features of real players' decisions and decision times. However, we also find that pretense was characterized by over-acting, stereotypical behavior, and suboptimal incorporation of new information, although these were undetected by human observers.

@@

```{r design, echo=FALSE, fig.cap="Experimental Design in Exp. 1 (upper panel) and 2 (lower panel). In non-pretend games, players revealed ships by guessing cells in a grid (A) or revealed a word by guessing letters (D). In pretend games, we marked ship locations with a cross (B) and revealed the target word from the start (E), but asked players to play as if they didn’t have this information. Lastly, players watched replays of the games of previous players and guessed which were pretend games (C and F).", out.width = '100%'}
knitr::include_graphics("figures/methods2.png")
```

@@

```{r sd_misses, echo=FALSE, cache=TRUE}

sd_misses_df <- E2.click_df %>%
  dplyr::select(subj_id,test_part, num_clicks) %>%
  mutate(subj_id=as.character(subj_id)) %>%
  # rbind(E2.optimal_click_df) %>%
  # rbind(E2.random_click_df) %>%
  mutate(test_part = factor(test_part, levels=c(
    'nonpretend','pretend'), labels = c(
    'non-pretend','pretend'))) %>%
  group_by(subj_id,test_part) %>%
  summarise(sd_misses=sd(num_clicks)) %>%
  mutate(game='Battleship') %>%
  rbind(
    E4.click_df %>%
  dplyr::select(subj_id,test_part, num_misses) %>%
  mutate(subj_id=as.character(subj_id)) %>%
  # rbind(E2.optimal_click_df) %>%
  # rbind(E2.random_click_df) %>%
  mutate(test_part = factor(test_part, levels=c(
    'nonpretend','pretend'), labels = c(
    'non-pretend','pretend'))) %>%
  group_by(subj_id,test_part) %>%
  summarise(sd_misses=sd(num_misses)) %>%
    mutate(game='Hangman')
  )

sd_misses_df_wide <- sd_misses_df %>%
  mutate(test_part=ifelse(test_part=='pretend','p','np'))%>%
  pivot_wider(values_from=sd_misses,
              names_from=c(test_part,game))

p <- sd_misses_df %>%
  mutate(sd_misses = round(sd_misses*2)/2)%>%
  ggplot(aes(x=sd_misses,fill=test_part, color=test_part))+
  scale_fill_manual(values=c(colors$np,colors$p, colors$random, colors$greedy)) +
  scale_x_continuous(limits=c(0,8), breaks=seq(0,10)) +
  labs(x='median number of misses', y='number of players')+
  theme(legend.position='none') +
  theme_classic() +
  # geom_vline(data=ref_lines, aes(xintercept = xintercept), color='black', linetype=2) +
  # geom_text(data=ref_lines, aes(x = xintercept, y=y, label=label), angle = 90, vjust = -1, size = 3, inherit.aes = FALSE)+
  geom_bar(alpha=0.6, position = 'identity', colour='black', size=0.5) +
  facet_wrap(~game, nrow=2) +
  theme(legend.position='na',strip.background = element_blank(), panel.border=element_blank())+
  labs(y='number of players',x="\u03C3(number of misses)")

ggsave('figures/sd_misses_facet.png',p, width=2.5, height=3)

## Stereotypical, imperfect self-simulation

Though impressive, the capacity for simulating a state of ignorance was not perfect. Importantly, the limitations and biases we observe are consistent with the simulation of a stereotypical, “cartoon” model of decision-making, rather than leakage of concealed information into the decision-making process. First, despite showing the same qualitative effects, decision time patterns in Battleship pretend games (but not Hangman pretend games) were systematically more pronounced relative to non-pretend games: a form of “over-acting”. Furthermore, pretend games followed stereotypical patterns, and as a result were more homogeneous than non-pretend games. Despite a highly similar average number of misses in pretend and non-pretend games (Fig. \@ref(fig:decisions)A), the number of unsuccessful guesses was overwhelmingly less variable in pretend relative to non-pretend games (Battleship: sd=`r sd_misses_df_wide$p_Battleship%>%mean(na.rm=T)%>%printnum()` in pretend versus `r sd_misses_df_wide$np_Battleship%>%mean(na.rm=T)%>%printnum()` in non-pretend games, $p$ `r apa_print(sd_misses_df_wide%>%mutate(diff=p_Battleship-np_Battleship)%>%pull(diff)%>%t.test())$table$p.value`; Hangman: sd=`r sd_misses_df_wide$p_Hangman%>%mean(na.rm=T)%>%printnum()` in pretend versus `r sd_misses_df_wide$np_Hangman%>%mean(na.rm=T)%>%printnum()` in non-pretend games, $p$ `r apa_print(sd_misses_df_wide%>%mutate(diff=p_Hangman-np_Hangman)%>%pull(diff)%>%t.test())$table$p.value`; Fig. \@ref(fig:rigidity)A).  Moreover, the within-participant sequential guess entropy, a measure of the inverse predictability of decision sequences across different games of the same player, was significantly reduced in pretend games (Fig. \@ref(fig:rigidity)B; see [online supplementary materials](https://matanmazor.github.io/ignorance/docs/exp.-1-battleship-exploratory-results.html#within-participant-click-entropy)), consistent with an attempt to enact typical, or average, behavior in a state of ignorance. Finally, Hangman pretenders were more likely to guess letters that appear frequently in English words (E, T, A, etc.) irrespective of the game state, compared to genuine players (Fig. \@ref(fig:rigidity)C). This suggests that in their attempt to behave as if they didn't know the true state of the game, pretenders had an increased tendency to follow rigid heuristics and rules, ignoring useful information as a result.

This limitation on incorporating evidence into the (simulated) decision-making process was especially evident in Hangman half-games, where players completed the game from a half-completed state. When asked to reveal the hidden fruit "\_A_A_A", `r round(100*E4.first_click_half_games %>% filter(test_part=='nonpretend_half_game' & word %in% c('banana', 'papaya'))%>%pull(bananahand)%>%mean())`% of the non-pretenders guessed one of the letters 'B' or 'N' (Fig. \@ref(fig:rigidity)D, yellow bars in left column). Among pretenders who knew that the hidden word was BANANA, this preference was reduced to `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word== 'banana')%>%pull(bananahand)%>%mean())`%. Importantly, half of the pretenders were given different information: they were told that the hidden word was the less prototypical fruit PAPAYA. Although good pretenders should simulate their behavior had they not known this information, only `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='papaya')%>%pull(bananahand)%>%mean())`% selected the letters 'B' or 'N', revealing that many pretenders were unable to predict that the first fruit that would have come to their minds was BANANA, not PAPAYA (Fig. \@ref(fig:rigidity)D, yellow bars in right column). A similar pattern was observed for the prototypical body part word HA(ND) and its surprising counterpart HA(IR): when playing normally, `r round(100*E4.first_click_half_games %>% filter(test_part=='nonpretend_half_game' & word %in% c('hand', 'hair'))%>%pull(bananahand)%>%mean())`% of the players selected letters that are consistent with the prototypical option HAND. This figure was `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='hand')%>%pull(bananahand)%>%mean())`% among pretenders for whom the target word was HAND, in contrast to only `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='hair')%>%pull(bananahand)%>%mean())`% among pretenders for whom the target word was HAIR (Fig. \@ref(fig:rigidity)D, blue bars).

```{r rigidity, echo=FALSE, message= F, warning=F, fig.cap="Limitations on flexible decision-making when pretending. A: variability in the number of misses (extracted individually for each player and then averaged) was lower in pretend games. B: sequential guess entropy, a measure of the (inverse) predictability of individual players’ guesses as a function of click number and guess number. In both Battleship and Hangman, sequential guess entropy increased with click number, and was overall lower in pretend games. C: letter frequency of Hangman guesses: the median rank frequency per participant in pretend and non-pretend games, with reference lines for the expected rank frequency for a random agent, and for a “greedy” agent that maximizes the probability of a hit in each step. D: first letter guesses in Hangman half-games, as a function of pretense condition and target word. Letters appear in alphabetical order. letters that appear in the target word are marked in yellow (BANANA), red (PAPAYA), blue (HAND) and magenta (HAIR). For reference, the overall distribution of letter guesses across all games and conditions is given below.", out.width = '100%'}
knitr::include_graphics("figures/stereotypical2.png")
```

@@

Despite these limitations, pretenders' behavior resembled real players' behavior closely enough that they were able to fool other participants into believing they were playing for real. After performing pretend and non-pretend games, participants were presented with game replays of previous players and took on the new role of being a 'judge' who had to determine who was pretending and who was playing for real. In both games, pretense detection was at chance (Battleship: `r E2.judge_df%>%group_by(subj_id)%>%summarise(accuracy=sum(correct)*20)%>%pull(accuracy)%>%mean()%>%round()`%, Hangman: `r round(E4.judge_summary_df$accuracy%>%mean()*100)`%). This is in line with previous findings of near-chance accuracy in lie detection [@bond2006]. Moreover, we find no sign of a correlation between pretense quality (measured as players' ability to trick judges into thinking they were not pretending) and pretense detection ability (measured as proportion correct; Battleship: `r apa_print(cor.test(E2.performance_df$judge_rate,E2.performance_df$pretend_rate, method='spearman'))$estimate`, Hangman: `r apa_print(cor.test(E4.performance_df$pretend_rate,E4.performance_df$judge_rate,method='spearman'))$estimate`), indicating that pretense and pretense detection rely on at least partly different cognitive processes.

@@

An alternative interpretation of our results is that instead of simulating a counterfactual knowledge state, participants actively suppressed or ignored the revealed game state such that their entire cognitive machinery was available to play the game. This would not require self-simulation, only a capacity to intentionally 'unsee', or forget, relevant evidence. While we cannot fully rule out this interpretation, we think it is unlikely to explain our players' successful pretense, for at least five reasons. First, we tried to make such suppression as hard as possible, by presenting the game solution on top of the game board for the entire duration of pretend games, and by having participants type the target word before pretend Hangman games. Second, suppressing thoughts on demand is notoriously difficult, and often has an opposite, positive effect on the suppressed content [@wegner1987paradoxical]. Third, when asked how they had performed the task in a debrief question, the responses of a significant majority of participants were aligned with self-simulation or rule-following (see [exploratory analysis](https://matanmazor.github.io/ignorance/docs/exp.-1-battleship-exploratory-results.html#suppression-or-simulation)). Fourth, response time patterns were exaggerated in pretend Battleship relative to non-pretend Battleship games (see Fig. \@ref(fig:time)A): a finding that is consistent with the simulation of a simplified self-model, but is hard to explain if participants were merely suppressing their current knowledge. Finally, pretend games were more similar to each other than were non-pretend games to themselves. This is again consistent with an attempt to simulate typical behavior.

@@

The research complied with all relevant ethical regulations and was approved by the Institutional Review Board of Johns Hopkins University. `E2.df$subj_id%>%unique()%>%length()` Participants were recruited via Prolific (prolific.co) and gave their informed consent prior to their participation. They were selected based on their acceptance rate (\>95%) and for being native English speakers. The entire experiment took approximately 20 minutes to complete. Participants' pay was equivalent to an hourly wage of 9.50 USD, in addition to a bonus payment (`r E2.bonus$bonus%>%min()` - `r E2.bonus$bonus%>%max()` USD, mean = `r E2.bonus$bonus%>%mean()%>%printnum()`).
